---
title: "Final Session — Solutions"
subtitle: "Statistical Tests · UpSet Plots · Custom Functions"
format:
  html:
    toc: true
---

```{r}
#| echo: false
#| warning: false
#| message: false
library(tidyverse)
library(ggplot2)
options(digits = 4)
```

---

## The Dataset: `iris`

```{r}
# Load and preview the iris dataset
data(iris)
glimpse(iris)

# Quick summary by species
iris |>
  group_by(Species) |>
  summarize(
    mean_sepal_length = mean(Sepal.Length),
    mean_petal_length = mean(Petal.Length),
    n                 = n()
  )
```

---

## Part 1: Statistical Tests

### 1a. Visualize the data first

```{r}
# Always plot before testing — it reveals the structure of your data
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.7, show.legend = FALSE) +
  geom_jitter(width = 0.15, alpha = 0.4) +
  labs(title = "Sepal Length by Species", y = "Sepal Length (cm)") +
  theme_minimal()
```

### 1b. Check normality within each species

```{r}
# Shapiro-Wilk test: H0 = data are normally distributed
# A non-significant p-value (p > 0.05) means we do NOT reject normality

shapiro.test(iris$Sepal.Length[iris$Species == "setosa"])
shapiro.test(iris$Sepal.Length[iris$Species == "versicolor"])
shapiro.test(iris$Sepal.Length[iris$Species == "virginica"])

# Q-Q plots for visual confirmation
par(mfrow = c(1, 3))
for (sp in levels(iris$Species)) {
  qqnorm(iris$Sepal.Length[iris$Species == sp], main = sp)
  qqline(iris$Sepal.Length[iris$Species == sp])
}
par(mfrow = c(1, 1))
```

### 1c. One-sample t-test

```{r}
# Question: Is the mean Sepal.Length of setosa equal to 5.0 cm?
# H0: mu = 5.0   H1: mu ≠ 5.0

setosa <- iris[iris$Species == "setosa", ]

result_one_sample <- t.test(setosa$Sepal.Length, mu = 5.0)
result_one_sample

# Interpretation:
# t-statistic, degrees of freedom (n-1 = 49), p-value, and 95% CI
# If p < 0.05 we reject H0 that the mean equals 5.0
```

### 1d. Two-sample t-test (Welch's, unequal variances)

```{r}
# Question: Is Sepal.Length different between setosa and versicolor?
# H0: mu_setosa = mu_versicolor

versicolor <- iris[iris$Species == "versicolor", ]

result_two_sample <- t.test(setosa$Sepal.Length,
                             versicolor$Sepal.Length)
result_two_sample

# Formula interface — cleaner syntax for subsetting
sub2 <- iris[iris$Species != "virginica", ]
t.test(Sepal.Length ~ Species, data = sub2)
```

```{r}
# Check equal variances first (F-test)
# H0: variances are equal
var.test(Sepal.Length ~ Species, data = sub2)

# If variances are equal (p > 0.05), can use var.equal = TRUE (Student's t)
t.test(Sepal.Length ~ Species, data = sub2, var.equal = TRUE)
```

### 1e. One-sample Z-test

```{r}
# Z-test: use when population sigma is known (rare in practice)
# For large n, t-test and Z-test give nearly identical results
# install.packages("BSDA") if needed
# BSDA stands for Basic Statistics and Data Analysis
library(BSDA)

# Assume we "know" sigma = 0.35 for setosa Sepal.Length
z.test(setosa$Sepal.Length,
       mu      = 5.0,
       sigma.x = 0.35)

# Compare p-value to the t-test result above — nearly identical for n=50
```

### 1f. Non-parametric alternative: Wilcoxon test

```{r}
# Use when normality cannot be assumed and n is small

# Two-sample Wilcoxon rank-sum (= Mann-Whitney U)
wilcox.test(Sepal.Length ~ Species, data = sub2)

# One-sample Wilcoxon signed-rank test
wilcox.test(setosa$Sepal.Length, mu = 5.0)
```

### 1g. One-way ANOVA — comparing all three species

```{r}
# Question: Does mean Petal.Length differ across all three species?
# H0: mu_setosa = mu_versicolor = mu_virginica

fit_anova <- aov(Petal.Length ~ Species, data = iris)
summary(fit_anova)

# A significant F-test means at least one pair differs.
# Use Tukey's HSD to find which pairs:
TukeyHSD(fit_anova)
```

```{r}
# Non-parametric alternative: Kruskal-Wallis
kruskal.test(Petal.Length ~ Species, data = iris)
```

### 1h. Correlation test

```{r}
# Question: Is there a linear relationship between Sepal.Length and Petal.Length?
# H0: rho = 0

# Pearson (assumes bivariate normality)
cor.test(iris$Sepal.Length, iris$Petal.Length, method = "pearson")

# Spearman (non-parametric, rank-based)
cor.test(iris$Sepal.Length, iris$Petal.Length, method = "spearman")

# Visualize the relationship
ggplot(iris, aes(x = Sepal.Length, y = Petal.Length, color = Species)) +
  geom_point(alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Sepal Length vs Petal Length") +
  theme_minimal()
```

### 1i. Multiple testing correction

```{r}
# Run t-tests for all three pairwise species comparisons
species_pairs <- list(
  c("setosa",     "versicolor"),
  c("setosa",     "virginica"),
  c("versicolor", "virginica")
)

p_values <- sapply(species_pairs, function(pair) {
  x1 <- iris$Sepal.Length[iris$Species == pair[1]]
  x2 <- iris$Sepal.Length[iris$Species == pair[2]]
  t.test(x1, x2)$p.value
})

names(p_values) <- sapply(species_pairs, paste, collapse = " vs ")
p_values

# Adjust for multiple comparisons
p.adjust(p_values, method = "bonferroni")   # conservative
p.adjust(p_values, method = "BH")           # FDR, less conservative
```

---

## Part 2: UpSet Plot

UpSet plots visualise intersections of multiple sets — a scalable alternative to Venn diagrams when you have more than 3 sets. Here we treat each iris measurement as a "set": a flower belongs to the set if its value exceeds the overall median.

```{r}
# install.packages("UpSetR")
library(UpSetR)

# Create a binary membership matrix:
# 1 = this flower is ABOVE the overall median for that measurement
# 0 = at or below the median

upset_data <- data.frame(
  long_sepal = as.integer(iris$Sepal.Length > median(iris$Sepal.Length)),
  wide_sepal = as.integer(iris$Sepal.Width  > median(iris$Sepal.Width)),
  long_petal = as.integer(iris$Petal.Length > median(iris$Petal.Length)),
  wide_petal = as.integer(iris$Petal.Width  > median(iris$Petal.Width))
)

head(upset_data)
```

```{r}
# Plot the UpSet diagram
upset(
  upset_data,
  sets           = c("long_sepal", "wide_sepal", "long_petal", "wide_petal"),
  order.by       = "freq",          # sort intersection bars by size
  main.bar.color = "steelblue",     # colour of intersection bars
  sets.bar.color = "darkorange",    # colour of set size bars (left)
  text.scale     = 1.3,
  mb.ratio       = c(0.6, 0.4)     # proportion: top bars vs dot matrix
)

# Reading the plot:
# - Left horizontal bars: total number of above-median flowers per measurement
# - Top vertical bars: size of each intersection (combination of traits)
# - Dot matrix: which measurements are involved in each intersection
# - Single dot = flowers above median in ONLY that measurement
# - Connected dots = flowers above median in ALL connected measurements
```

---

## Part 3: Custom Functions

### 3a. Basic function syntax

```{r}
# A function to compute the standard error of a numeric vector
std_error <- function(x) {
  sd(x) / sqrt(length(x))
}

std_error(iris$Sepal.Length)
std_error(iris$Petal.Length)
```

### 3b. Functions with default arguments

```{r}
# Summarise one measurement for one species
# 'variable' has a default so it's optional
summarise_species <- function(data, species_name,
                               variable = "Sepal.Length") {
  sub <- data[data$Species == species_name, variable]

  list(
    mean = mean(sub),
    sd   = sd(sub),
    se   = sd(sub) / sqrt(length(sub)),
    n    = length(sub)
  )
}

# Using the default variable
summarise_species(iris, "setosa")

# Overriding the default
summarise_species(iris, "virginica", "Petal.Length")
```

### 3c. Applying a function across a list with `sapply` and `lapply`

```{r}
# Compute mean Petal.Length for every species without copy-pasting
species_list <- levels(iris$Species)

# sapply: simplifies to a vector/matrix where possible
mean_petals <- sapply(species_list, function(sp) {
  mean(iris$Petal.Length[iris$Species == sp])
})
mean_petals

# lapply: always returns a list — useful when output is complex
summaries <- lapply(species_list, function(sp) {
  summarise_species(iris, sp, "Petal.Length")
})
names(summaries) <- species_list
summaries[["setosa"]]
```

### 3d. A plotting function — reuse across measurements

```{r}
# Function to make a boxplot + jitter for any measurement
plot_by_species <- function(data, variable, fill_colors = NULL) {

  if (is.null(fill_colors))
    fill_colors <- c("setosa"     = "#619CFF",
                     "versicolor" = "#F8766D",
                     "virginica"  = "#00BA38")

  ggplot(data, aes(x = Species,
                   y = .data[[variable]],
                   fill = Species)) +
    geom_boxplot(alpha = 0.6, show.legend = FALSE) +
    geom_jitter(width = 0.15, alpha = 0.4, show.legend = FALSE) +
    scale_fill_manual(values = fill_colors) +
    labs(title = paste(variable, "by Species"),
         y = variable, x = NULL) +
    theme_minimal()
}

# Call once per measurement — same function, different data
plot_by_species(iris, "Sepal.Length")
plot_by_species(iris, "Petal.Length")
plot_by_species(iris, "Petal.Width")
```

### 3e. Putting it all together — a mini analysis pipeline

```{r}
# A single function that runs the full pairwise comparison for one variable
pairwise_analysis <- function(data, variable = "Sepal.Length") {

  cat("=== Pairwise t-tests for", variable, "===\n\n")

  pairs <- combn(levels(data$Species), 2, simplify = FALSE)

  p_vals <- sapply(pairs, function(pair) {
    x1 <- data[data$Species == pair[1], variable]
    x2 <- data[data$Species == pair[2], variable]
    t.test(x1, x2)$p.value
  })

  names(p_vals) <- sapply(pairs, paste, collapse = " vs ")

  result <- data.frame(
    comparison  = names(p_vals),
    p_value     = round(p_vals, 4),
    p_bonf      = round(p.adjust(p_vals, "bonferroni"), 4),
    p_BH        = round(p.adjust(p_vals, "BH"), 4),
    significant = p.adjust(p_vals, "BH") < 0.05
  )

  print(result)
  invisible(result)
}

pairwise_analysis(iris, "Sepal.Length")
pairwise_analysis(iris, "Petal.Length")
```
